Venn diagrams indicate erroneous analyses in transcriptomics

### 2022-01-06 mk4/response_to_editor.md

**We fully agree that FNs are the core of the problem! However, we know
from other studies (and basic statistics) that even in most favorable
circumstances the FNR will likely be above 20%; and that in case of transcriptomic
studies, the FNR may be much larger (please see the relevant fragment already
included in the paper). In other words, it does not matter whether one
looks at 80 samples or 193 samples: FNs are bound to occur. That is the
reason – we claim – for the fact that the gene set enrichments found in
this manner are related to the studied problem. However, we have addressed
this isssue by testing different sample sizes (see below).**

Therefore,
the design of the experiment is biased in favour of the "desired" conclusion.

**We have mimicked an RNA-Seq experiment using a sample size
well above the median used by others precisely to avoid this allegation. We
have now included a relevant reference (Baccarella et al.) which includes a
literature survey on RNA sample size in real world data. In there you will
see that half of the studies used sample sizes sizes of 6 or less per
group.

Nonetheless, we think that showing how the artifacts depend on sample size
is, in itself, interesting, not least because it shows that sample size has
a direct effect on the number of observed artifacts: the larger the sample
size, the more artifacts are observed. We have now included a section
comparing the results for 10, 20 (our example) and 40 samples per
group/treatment combination (thus, total sample sizes of 40, 80 and 160).
The reason for this is, of course, that the more DEGs you observe, the more
powerful the hypergeometric test is.

@@

For each sample size, we have run 100 replicates of the group sampling,
replicated the whole procedure and gathered all test results. Results are
shown on the new Fig. 3, C and D.

However, we also think that the core of the problem is the incorrect
statistical procedure (selecting genes by comparing significance with lack
thereof) which will not become more correct by increasing the sample size.**

@@

The authors use a small log-fold change value of 1 to define the
differential expression. This leads to a high number of false positives.

**Statistical theory (e.g. the above reference to the original Benjamini &
Hochberg paper) tells us that the false discovery rate in differential
expression analysis is controlled by the BH procedure, not by the log fold
change threshold, except incidentally (beause log fold change is correlated
to the p-value, thus setting a more conservative log2 fold change results
in eliminating most of the larger p-values). Genes with large fold changes
may still not be significant and vice versa, true positives may have small
log2 fold changes. Therefore you cannot predict what will be the precise
effect of raising the LFC threshold on FDR except in very broad terms,
whereas setting the FDR threshold gives you clear information about the
expected proportion of false positives in your results.

This is analogous to using an effect size filter (say, Cohen's d) on
results of a t-test. The p-value from the t-test controls the type I error
rate, but a statistically significant difference does not necessarily
indicate a biologically meaningfull effect. For this, we turn to the effect
size (see for example Sullivan, Gail M., and Richard Feinn. "Using effect
size—or why the P value is not enough." Journal of graduate medical
education 4.3 (2012): 279-282.).

Thus, the threshold serves to eliminate genes that do show statistically
significant change (even if they are true positives), but which are not
biologically relevant (because the effect size, for which log2 fold change
is a proxy, is too small). The proportion of false positives is 5% even with no
log2 FC treshold.

Increasing the threshold can, of course, result in not observing any
enrichments (see Fig. 3, A and B). Indeed, it does for some replicates.
This is not surprising – without a sufficient number of DEGs, no
enrichments can be calculated using a hypergeometric test. However, even
though the effect of increasing log2 fold change may sometimes eliminate the
symptoms, it does not cure the disease – the core problem of erroneous
analysis of interactions remains.

Nonetheless, we have compared the effect of the log2
fold change threshold on the results (Fig. 3).**

@@

Higher values of log-fold change should also be used, and the results
should be compared and discussed.

**We have now included a section in the manuscript which shows the
dependence of the artifacts on the log2 fold change thresholds used.
For this, we have replicated the whole procedure of group randomization 100
times. Summary overview can be found in the new Fig. 3. Bottom line: yes,
for many replicates the artifacts disappear when the log2 fold change threshold is high –
because the DEGs disappear. However, this is not generally so: even for LFC
greater or equal 3 and total sample size of 160 the artifacts are still
present in 78 out of 100 replicates.**

### 2022-01-06 presentation/manuscript.rmd
added presentation

---
title: "Venn diagrams may indicate erroneous statistical approaches leading to artifacts in transcriptomic analyses"
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
output:
  html_document
outputoff:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 20 # 2 strata per condition 
```

@@

set.seed(20220105)

@@

ds2_file <- "ds2_cached.rds"
if(!file.exists(ds2_file)) {
  message("Running DESeq2")
  ds2 <- DESeq(ds2)
  saveRDS(ds2, file=ds2_file)
} else {
  message("Reading ds2 from manual cache")
  ds2 <- readRDS(ds2_file)
}
```

@@

```{r GSEA}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
```

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. To investigate whether the sampling had an impact on
the artifacts, we have repeated the above procedure for 100 replicates,
each containing a different set of samples randomly assigned to the two
groups.  Furthermore, we tested whether the selection of the log~2~ fold
change threshold
or different sample size might have an impact on the extent of the arising
artifacts. To this end, we have tested `r length(lfc_thresholds)` different
threshold values for the log~2~ fold changes and three sample sizes: 20, 40 or 80 samples
per group (corresponding to 10, 20 or 40 samples per group/treatment
combination).

**Fig. 3.Influence of log~2~ fold change treshold (A, C) and the sample
size (B, D) on the number of genes significant in one, but not in the other
group (A, B) and the number of significant gene sets found in one group, but
not the other (C, D).**  Top row shows the influence of log~2~ fold change
threshold for sample size of 40 per group (20 per group/treatment
combination), bottom row shows the influence of sample size for a log~2~
fold change threshold set to `r lfc_thr`.

```{r montecarlo,cache=TRUE,fig.width=5.5,fig.height=5.5}
## note: given that the monte carlo simulation is quite time consuming, we
## have included the code for it in another file, `montecarlo.R`. If trying
## to replicate this manuscript, please execute that file first to generate
## the necessary replicates.

@@

>>>>

---
title: ""
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
output:
  html_document
outputoff:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per condition 
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```

@@

set.seed(0123)

@@

ds2_file <- "ds2_cached.rds"
if(TRUE || !file.exists(ds2_file)) {
  message("Running DESeq2")
  ds2 <- DESeq(ds2)
  saveRDS(ds2, file=ds2_file)
} else {
  message("Reading ds2 from manual cache")
  ds2 <- readRDS(ds2_file)
}
```

@@

```{r GSEA,cache=FALSE}
common <- res.merged %>% filter(DEG.g1 & DEG.g2) %>% pull(SYMBOL)
gsea_res <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset)
})
gsea_res_full <- map(res, ~ {
  fg <- .x %>% filter(padj < padj_thr & abs(log2FoldChange) > lfc_thr)
  fg <- setdiff(fg$SYMBOL, common) # only "specific" genes
  tmodHGtest(fg=fg, bg=.x$SYMBOL, mset=mset, qval = Inf)
})
message("G1 ", nrow(gsea_res$g1), " G2 ", nrow(gsea_res$g2))
gsea_res.merged <- merge(gsea_res$g1, gsea_res$g2, by=c("ID", "Title"), suffixes=c(".g1", ".g2"), all=T)
saveRDS(gsea_res.merged, file="gsea_res_merged.rds")
```

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. Arguably, the observations might differ if the groups
were to be resampled. To investigate whether the sampling had an impact on
the artifacts, we have repeated the above procedure for 100 replicates,
each containing a different set of samples randomly assigned to the two
groups.  Furthermore, we tested whether the selection of the log~2~ fold
change threshold or different sample size might have an impact on the
extent of the arising artifacts. To this end, we have tested `r
length(lfc_thresholds)` different threshold values for the log~2~ fold
changes and three sample sizes: 20, 40 or 80 samples per group
(corresponding to 10, 20 or 40 samples per group/treatment combination).

**Fig. 3. Influence of log~2~ fold change treshold (A, C) and the sample
size (B, D) on the number of genes significant in one, but not in the other
group (A, B) and the number of significant gene sets found in one group, but
not the other (C, D).**  Top row shows the influence of log~2~ fold change
threshold for sample size of 40 per group (20 per group/treatment
combination), bottom row shows the influence of sample size for a log~2~
fold change threshold set to `r lfc_thr`.

```{r montecarlo,cache=TRUE,fig.width=7.5,fig.height=6.5}
## note: given that the monte carlo simulation is quite time consuming, we
## have included the code for it in another file, `montecarlo.R`. If trying
## to replicate this manuscript, please execute that file first to generate
## the necessary replicates.

@@

Setting a higher log~2~ fold change threshold reduces the number of
DEGs and of observed artifacts (Fig. 3, A and B). However, even for log~2~
threshold of 3 (DEGs defined by 8-fold change and FDR < `r pval_thr`) the
number of replicates in which artifacts can be observed is 
`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
(out of 100),
and in at least 
`r sum(gsea_by_lfc_thr %>% filter(lfc == 3) %>% pull(value) > 0)`
replicates, 5 or more gene sets were significantly
enriched. Thus, setting a more conservative threshold cannot protect from
the arising artifacts.

Contrary to our expectations, the number of artifacts rose with sample size
(Fig. 3, C and D). For total sample size of 160 (40 samples per
group/treatment combination) the number of artifacts was almost 
`r round(
  mean(gsea_by_ss %>% filter(ss == 40) %>% pull(value)) /
  mean(gsea_by_ss %>% filter(ss == 20) %>% pull(value)))
`
times higher than for total sample size of 80.
